\bf Excuse me, do you have a moment to talk about version control?

### 2017-05-19 README.tex
Make both formats play nice re: bibliography

---
title: "Excuse me, do you have a moment to talk about version control?"
authors:
- name: Jennifer Bryan
  affiliation: |
    RStudio and the Department of Statistics, University of British Columbia
  thanks: |
    The author gratefully acknowledges the constructive feedback from reviewers
    Nicholas Horton and Colin Rundel.
blinded: 0
output:
  rticles::asa_article: null
  github_document:
    toc: yes
keywords: version control
abstract: Abstract abstract abstract.
bibliography: git-github-for-stats.bib
---

```{r include = FALSE}
## I temporarily embed some figures from the internet and LaTeX just. can't.
## deal. with. it.
is_github_document <-
  knitr::opts_knit$get("rmarkdown.pandoc.to") == "markdown_github"
```

## Why Git?

@@

  * Doing your work becomes tightly integrated with organizing, recording, and disseminating it. It's not a separate, burdensome task you are tempted to neglect.
  * Collaboration is much more structured, with powerful tools for asynchronous work and managing versions.
  * The marginal effort required to create a web presence for a project is negligible.
  * GitHub makes a fantastic course management system for courses that use R. You and your students can exchange actual working code and explore the associated results.
  * By using common mechanics across work modes (research, teaching, analysis), you achieve basic competence quickly and avoid the demoralizing forget-relearn cycle.

## What is Git?

[Git](http://git-scm.com) is a __version control system__. Its original purpose was to help groups of developers work collaboratively on big software projects. Git manages the evolution of a set of files -- called a __repository__ or __repo__ -- in a sane, highly structured way. It is a bit like the "Track Changes" feature from Microsoft Word, but more rigorous, powerful, and scaled up to multiple files.

@@

In my opinion, **for new users**, the pros of Git only outweigh the cons when you factor in the overhead of communicating and collaborating with other people, including your future self. And who among us does not need to do that? Your life is much easier if this is baked into your workflow, as opposed to being a separate process that you dread and avoid. Communication and collaboration are the killer apps of version control. Git enforces a rigorous model of file management, but it is critical for the distribution of files across different people, computers, and time.

@@

Many people who don't use Git unwittingly re-invent a poor man's version of it. They take an important file and distribute it via email. Various parties make changes, decorating the file name with initials, dates, and other descriptors. Before you know it, the original file is the root of a complicated phylogeny that no amount of "Track changes" and good intentions can resolve.

*If you're looking at the PDF draft, you're missing out on some figures around here in the GitHub Markdown version that hint at the figure I plan to embed.*

```{r eval = is_github_document, echo = FALSE, results = 'asis'}
cat("*here's one way I see phylogeny in my mind*\n")
cat("![](https://upload.wikimedia.org/wikipedia/commons/2/2b/Phylogeny_of_Retroviruses.jpg)")

cat("*make such a phylogeny with a set of crazy filenames like this*\n")
cat("![](https://www.onshape.com/hs-fs/hub/379003/file-2533844795-jpg/Blog_Assets/Version_Control_Graphic-2.jpg?t=1494963046029)")

cat("*or riff on the famous PhD Comic*")
cat("![](http://phdcomics.com/comics/archive/phd101212s.gif)")
```

@@

*another good image to emulate, from the Bartlett Git for Designers talk*
![](bartlett-copies-vs-commits.png)

## Who should read this and what to expect

@@

## What is GitHub?

[GitHub](https://github.com) is currently the most popular Git hosting service. Others include [GitLab](https://about.gitlab.com) and [Bitbucket](https://bitbucket.org). These services provide a home for Git-based projects on the internet. It is a bit like DropBox or Google Drive, but more structured, powerful, and programmatic.

@@

  1. When you are new with Git (or, frankly, even when you're not), it's common to goof up the Git infrastructure for a project. Note that your files can absolutely be intact and safe, even while the Git tracking is a bit confused. Of course there are official Git remedies, but sometimes the easiest fix is to clone a fresh copy from GitHub, patch things up with the changes that only exist locally, and move on with your life. This workaround obviously requires the existence of a recent copy on GitHub.
  2. The highly functional web interfaces mentioned above are often the most pleasant and natural way to navigate and search your files, even though all the same information exists locally. It is a pleasure to browse through your own work, across multiple projects or files and across time, as if it's a well-designed website. You must push your work to GitHub to enjoy this.

GitHub __issues__ are another powerful feature of the platform. Recall that we are repurposing Git, a tool designed to facilitate software development. The issues for a project are its bug tracker. For projects that are not pure software development, we co-opt this machinery to organize our to do list more generally. The basic unit is an issue and you can interact with them in two ways.

@@

Issues can be assigned to specific people and they can be labelled, e.g. "bug", "simulation-study", or "final-exam". Coupled with the ability to cross-link issues and the project files or file changes, you have extraordinary power to document why things have happened in the past and to organize what needs to happen in the future.

## Initial system setup

@@

  * Register for a free account with GitHub.

  * Install Git. Depending on your OS, Git might already be installed. But many of us will need to install it or might choose to update to the most recent version. Some basic configuration is critical, such as setting your username and email.

  * Install a local Git client. Optional but highly recommended. A Git client is software that provides a graphical user interface for Git, which is otherwise command-line only. If you are an R user, you will find that RStudio provides a great deal of this functionality. There are some notable gaps, however, so you might still choose to install a dedicated and comprehensive Git client such as SourceTree or GitKraken. Git is a file-based system, so you can do some operations from RStudio, others from SourceTree, and others from the shell.

  * Confirm, with a practice repository, that local Git can send and receive the current version of the repository on GitHub, known as *pushing* and *pulling*, respectively. This will require authenticating yourself with GitHub from a local shell or Git client. At this point, most people elect to do a bit of extra setup to ensure that they are not repeatedly challenged for their GitHub credentials going forward.

@@

## Is this going to hurt?

@@

  * I repeat: consider using a graphical front-end for Git, a.k.a. a "Git client", versus restricting yourself to the command line interface.
  * Establish confidence in the basics (e.g. make a change, commit it, push it) before wading into more advanced usage (e.g. branching).
  * Commit yourself to Git usage on a project that will provide sustained practice over several months. Usage in a course is great, because it provides a relentless stream of small deadlines.
  * Realize that no one is giving out Git style points. It's ok to "power-cycle", i.e. re-initialize the Git repository, to get unstuck.

## Repositories and workflow

@@

  * Dedicate a local directory (a.k.a "folder") to it.
  * Make it an RStudio Project. Optional but recommended; obviously only applies to projects involving R and users of RStudio.
  * Make it a Git repository.

@@

  * Go about your usual business, e.g. writing R scripts or authoring reports in LaTeX or R Markdown. But instead of only *saving* individual files, periodically you make a **commit**, which takes a snapshot of all the files in the entire project.
    - Have you ever versioned a file [by adding your initials or the date](http://www.phdcomics.com/comics/archive.php?comicid=1531)? That is effectively a **commit**, albeit only for a single file: it is a version that is significant to you and that you might want to inspect or revert to later.
  * Push commits to GitHub periodically.
    - This is like sharing a document with colleagues on DropBox or sending it out as an email attachment. By pushing to GitHub, you make your work and all your accumulated progress accessible to others.

This is a moderate change to your normal, daily workflow. It feels weird at first, but quickly becomes second nature. In [STAT 545](http://stat545.com) students are required to submit all coursework via GitHub, starting in week one. Most have never seen Git before and do not identify as "programmers". It is a major topic in class and office hours for the first two weeks. Then we practically never discuss it again.

*I am not saying anything about pulling and collaboration here, which seems a bit extreme, even if I can't say much*

## Commits, diffs, and tags

@@

  * repository
  * commit
  * diff

Recall that a repository -- or "repo" -- is just a directory of files that Git is going to manage holistically. A commit functions like a snapshot of all the files in the repo, at a specific moment. Under the hood, that is not exactly how Git implements things. Mental models don't have to be accurate in order to be useful, but in this case there's some value in aligning the two.

*this is really crying out for an example and/or diagrams, something that shows commits unfolding and let's you illustrate a diff, sort of like this from the barlett talk*

![](bartlett-commit-history.png)

Consider version A of a file and a modified version, version B. Assume that version A was part of a Git commit and version B was part of the next commit. The set of differences between A and B is called a "diff" and Git users contemplate diffs a lot. Diff inspection is how you re-explain to yourself how version A differs from version B. Diff inspection is not limited to adjacent commits. You can inspect the diffs between any two commits.

@@

Every time you make a commit you must also write a short *commit message*. Ideally, this conveys the *motivation* for the change. Remember, the diff will show the content. When you revisit a project after a break or need to digest recent changes made by a colleague, looking at the *history*, by reading commit messages and skimming through diffs, is an extremely efficient way to get up to speed.

Every commit needs some sort of nickname, so you can identify it. Git does this automatically, assigning each commit what is called a *SHA*, a seemingly random string of 40 letters and numbers (it is not, in fact, random but is a SHA-1 checksum hash of the commit). Though you will be exposed to these, you don't have to handle them directly very often and, when you do, usually the first 7 characters suffice. You can also designate certain snapshots as special with a *tag*, which is a name of your choosing. In a software project, it is typical to tag a release with its version, e.g., "v1.0.3". For a manuscript or analytical project, you might tag the version submitted to a journal or transmitted to external collaborators.

## Markdown is special on GitHub

@@

Any file written in Markdown is rendered in an HTML-like way on GitHub. In particular, formatting and links "just work". This is the last piece we need to seal my claim that merely pushing your project to GitHub gives it a web presence for zero extra work. If you make even a modest effort to embed a few explanatory Markdown files in your repo, you will get an automatically-updated project website for free. In particular, if a directory has a `README.md` file, GitHub renders it like a home page or "index.html" when people visit that directory in the browser. It is very common for a repo to have a top-level `README.md`, but each subdirectory can have its own as well.

## Markdown is special for R users

Markdown is extra special for R users because of R Markdown, which is just Markdown that includes chunks of R code. Again, do not regard R Markdown as   something you must clear your schedule to learn. If you can write email and a bit of R code, you can write R Markdown. The rmarkdown package converts R Markdown (`.Rmd` files) to Markdown (`.md` files), running the code and inserting the results, including figures, into the document. This is powered by another package, knitr, under the hood. This process is made especially easy in RStudio, but is by no means limited to users of that application.

*this too is crying out for a visual Rmd -> md -> rendered thing looking good*

These R-derived Markdown files, if committed and pushed, then enjoy the usual privileged treatment on GitHub already described above. Once an `.Rmd` file has been rendered to `.md`, anyone viewing it on GitHub can read the prose, study the R code, **and view the results of running that code**, including figures. It is the best of all worlds, because the code is revealed and, by definition, is the code that produced the results. And yet a reader can gaze upon the product in a web browser, without needing to download the code, install all necessary dependencies, and run it.

@@

You do not, in fact, even need to work in R Markdown to exploit this. It works with plain R scripts as well. You can use exactly the same machinery to prepare a rendered version of an R script, i.e. to go from `.R` to `.md`. Again, RStudio makes this especially easy, but this is not limited to RStudio. Once the Markdown file is pushed to GitHub, it is as if the reader has run your code or is able to look over your shoulder at your R session. This provides an lightweight system for exposing work-in-progress to collaborators, without slowing down to create separate reports. Comment lines that begin with `#'` are elevated to top-level prose, providing a way to make the document more welcoming for a reader. Once there are many prose comments, you might decide to switch from `.R` to `.Rmd`, have proper top-level prose, and move the code down into chunks.

*figure showing the yaml needed for this? so the use of `github_document` or `keep_md = TRUE`; try to make a figure that does double or triple duty, i.e. illustrates this and points made elsewhere*

Let's zoom back out again and consider R Markdown and the rmarkdown package more generally. I want to point out that R Markdown can be rendered to many more formats than Markdown! I have emphasized the production of Markdown here because it is extremely useful in GitHub-hosted projects and that seems to be underappreciated. But the rmarkdown package can convert `.Rmd` to a wide array of output formats, including HTML, PDF, and Word (`.docx`). The Markdown must be created regardless. Sometimes it is all you need, such as on GitHub. In other contexts, it is just a necessary intermediate and may even be discarded.

## Which files to commit

@@

  * R markdown `.Rmd` --> markdown `.md`
  * R markdown `.Rmd` --> markdown `.md` --> `.html` or `.pdf` or `.docx`
  * R script `.R` --> results as `.csv` or `.rds` and figures `.png`
  * LaTeX `.tex` --> WHAT GOES HERE NOW? --> `.pdf`

The files at the far left are clearly source files. In the case of an R script, this is literally true, but it's morally true for R markdown and LaTeX files too. These are files that you directly create and edit "by hand".

The files at the far right are clearly derived and are often described as "targets". These files are programmatically generated from source files (and possibly other inputs). These files are the product and they have external value, often for communicating ideas and results.

The files in the middle are intermediates. Like targets, they are programmatically generated, but, unlike targets, no one necessarily cares about them. However, note that intermediate Markdown `.md` is an exception, since it is extremely useful on GitHub -- much more so than `.html`, `.pdf`, or `.docx` -- and is more like an additional target.

A critical issue for workflow happiness is figuring out how to manage the production and storage of source, intermediates, and targets with respect to Git. You can direct Git to ignore specific files or certain types of files, such as autosaves created by your editor. This reduces clutter in your project: Git will not pester you to commit changes to these files and they will not appear in the associated GitHub repository. A file that Git does not ignore is said to be *tracked*.

@@

*Is it immediately useful to someone?* If so, that is a reason to track it and push it to GitHub. There is a taboo against committing derived products, inherited from Git's software development roots. The reasoning is that compiled programs are platform-specific and, therefore, people are better served by getting current source from Git and compiling themselves. I think data analytic targets, like figures and rendered reports, are very different beasts and it's misguided to reflexively exclude them from version control. They are immediately useful, especially to consumers of a project (versus the makers). To the extent that a GitHub repo is meant for dissemination, there is no reason to burden every consumer with unnecessary friction. Most simply will not bother to clone the repo, install all the necessary dependencies, and remake the products. Make them readily available.

*Is it available elsewhere?* If so, then perhaps you don't need to track it and push it to GitHub. Many people who have a policy of not tracking derived products also have a system that makes these available elsewhere, such as on a separate website. There are ways to automate this via GitHub, but that is beyond the scope of this article and not recommended for your early days with Git and GitHub. Beware those who recommend the exclusion of derived products without offering any practical solution for making them available some other way.

*Is it huge or changing often? Is it a format that is of little use on GitHub?* These are all good reasons to not track a file with Git. They can make your repository bloated and slow down pushes and pulls. If a file is binary, such as a Word document or Excel spreadsheet, Git and Git clients will not be able to provide human-readable diffs. Neither will GitHub be able to directly display this file in the browser. Be aware, however, GitHub-friendliness does not just boil down to "is it plain text?". GitHub has excellent support for non-code files, such as image formats (PNG, JPG, GIF, and SVG) and PDFs. It even provides visual diffs, which are extremely useful for understanding changes in figures. Finally, even though HTML is plain text, it is of little direct use on GitHub, because, unlike Markdown, it is not rendered.

*Will diffs be useful to you?* Some derived files are simply too big or miserable to read casually, such as `.csv` files of processed results or `.html` derived from `.Rmd`. But they may still be worth tracking with Git, because the diffs are often modest and quite interesting. I have caught unexpected changes in analytical results and student-facing webpages this way. When you re-run an analysis with updated input data or after updating R packages, the diffs presented by Git help you quickly pinpoint the downstream consequences.

*Will it make collaboration harder?* *talk about the scenario Nick brought up re: binary files like PDF being a common source of merge conflict <https://github.com/dsscollection/git-github-for-stats/issues/6>*

@@

## Collaboration

@@

  * **Edit, save, attach.** In this workflow, everyone has one (or more!) copies of the document and they circulate via email attachment, accumulating initials and dates in the filename. Which one is "master"? Does this question even make sense anymore? How do different versions of the document relate to each other? If you want to see a version combining the best versions of each section, how would you reconcile the different copies into one?  All of this usually gets sorted out by social contract, a fairly manual process, and at least one miserable person.

  * **Google Doc.** In this workflow, there is only one copy of the document and it lives in the cloud. Anyone can access the most recent version on demand. Anyone can edit or comment or propose a change and this is immediately available to everyone else. Anyone can see who's been editing the document and how and, if disaster strikes, can revert to a previous version. A great deal of ambiguity and annoying reconciliation work has been designed away.

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{listings}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Excuse me, do you have a moment to talk about version control?}

  \author{
        Jennifer Bryan \thanks{The author gratefully acknowledges the constructive feedback from
reviewers Nicholas Horton and Colin Rundel.} \\
    RStudio and the Department of Statistics, University of British Columbia\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Excuse me, do you have a moment to talk about version control?}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Abstract abstract abstract.
\end{abstract}

\noindent%
{\it Keywords:} version control
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\subsection{Why Git?}\label{why-git}

@@

\begin{itemize}
\tightlist
\item
  Doing your work becomes tightly integrated with organizing, recording,
  and disseminating it. It's not a separate, burdensome task you are
  tempted to neglect.
\item
  Collaboration is much more structured, with powerful tools for
  asynchronous work and managing versions.
\item
  The marginal effort required to create a web presence for a project is
  negligible.
\item
  GitHub makes a fantastic course management system for courses that use
  R. You and your students can exchange actual working code and explore
  the associated results.
\item
  By using common mechanics across work modes (research, teaching,
  analysis), you achieve basic competence quickly and avoid the
  demoralizing forget-relearn cycle.
\end{itemize}

\subsection{What is Git?}\label{what-is-git}

\href{http://git-scm.com}{Git} is a \textbf{version control system}. Its
original purpose was to help groups of developers work collaboratively
on big software projects. Git manages the evolution of a set of files --
called a \textbf{repository} or \textbf{repo} -- in a sane, highly
structured way. It is a bit like the ``Track Changes'' feature from
Microsoft Word, but more rigorous, powerful, and scaled up to multiple
files.

@@

In my opinion, \textbf{for new users}, the pros of Git only outweigh the
cons when you factor in the overhead of communicating and collaborating
with other people, including your future self. And who among us does not
need to do that? Your life is much easier if this is baked into your
workflow, as opposed to being a separate process that you dread and
avoid. Communication and collaboration are the killer apps of version
control. Git enforces a rigorous model of file management, but it is
critical for the distribution of files across different people,
computers, and time.

@@

Many people who don't use Git unwittingly re-invent a poor man's version
of it. They take an important file and distribute it via email. Various
parties make changes, decorating the file name with initials, dates, and
other descriptors. Before you know it, the original file is the root of
a complicated phylogeny that no amount of ``Track changes'' and good
intentions can resolve.

\emph{If you're looking at the PDF draft, you're missing out on some
figures around here in the GitHub Markdown version that hint at the
figure I plan to embed.}

@@

\emph{another good image to emulate, from the Bartlett Git for Designers
talk} \includegraphics{bartlett-copies-vs-commits.png}

\subsection{Who should read this and what to
expect}\label{who-should-read-this-and-what-to-expect}

@@

\subsection{What is GitHub?}\label{what-is-github}

\href{https://github.com}{GitHub} is currently the most popular Git
hosting service. Others include \href{https://about.gitlab.com}{GitLab}
and \href{https://bitbucket.org}{Bitbucket}. These services provide a
home for Git-based projects on the internet. It is a bit like DropBox or
Google Drive, but more structured, powerful, and programmatic.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  When you are new with Git (or, frankly, even when you're not), it's
  common to goof up the Git infrastructure for a project. Note that your
  files can absolutely be intact and safe, even while the Git tracking
  is a bit confused. Of course there are official Git remedies, but
  sometimes the easiest fix is to clone a fresh copy from GitHub, patch
  things up with the changes that only exist locally, and move on with
  your life. This workaround obviously requires the existence of a
  recent copy on GitHub.
\item
  The highly functional web interfaces mentioned above are often the
  most pleasant and natural way to navigate and search your files, even
  though all the same information exists locally. It is a pleasure to
  browse through your own work, across multiple projects or files and
  across time, as if it's a well-designed website. You must push your
  work to GitHub to enjoy this.
\end{enumerate}

GitHub \textbf{issues} are another powerful feature of the platform.
Recall that we are repurposing Git, a tool designed to facilitate
software development. The issues for a project are its bug tracker. For
projects that are not pure software development, we co-opt this
machinery to organize our to do list more generally. The basic unit is
an issue and you can interact with them in two ways.

@@

Issues can be assigned to specific people and they can be labelled, e.g.
``bug'', ``simulation-study'', or ``final-exam''. Coupled with the
ability to cross-link issues and the project files or file changes, you
have extraordinary power to document why things have happened in the
past and to organize what needs to happen in the future.

\subsection{Initial system setup}\label{initial-system-setup}

@@

\begin{itemize}
\item
  Register for a free account with GitHub.
\item
  Install Git. Depending on your OS, Git might already be installed. But
  many of us will need to install it or might choose to update to the
  most recent version. Some basic configuration is critical, such as
  setting your username and email.
\item
  Install a local Git client. Optional but highly recommended. A Git
  client is software that provides a graphical user interface for Git,
  which is otherwise command-line only. If you are an R user, you will
  find that RStudio provides a great deal of this functionality. There
  are some notable gaps, however, so you might still choose to install a
  dedicated and comprehensive Git client such as SourceTree or
  GitKraken. Git is a file-based system, so you can do some operations
  from RStudio, others from SourceTree, and others from the shell.
\item
  Confirm, with a practice repository, that local Git can send and
  receive the current version of the repository on GitHub, known as
  \emph{pushing} and \emph{pulling}, respectively. This will require
  authenticating yourself with GitHub from a local shell or Git client.
  At this point, most people elect to do a bit of extra setup to ensure
  that they are not repeatedly challenged for their GitHub credentials
  going forward.
\end{itemize}

@@

\subsection{Is this going to hurt?}\label{is-this-going-to-hurt}

@@

\begin{itemize}
\tightlist
\item
  I repeat: consider using a graphical front-end for Git, a.k.a. a ``Git
  client'', versus restricting yourself to the command line interface.
\item
  Establish confidence in the basics (e.g.~make a change, commit it,
  push it) before wading into more advanced usage (e.g.~branching).
\item
  Commit yourself to Git usage on a project that will provide sustained
  practice over several months. Usage in a course is great, because it
  provides a relentless stream of small deadlines.
\item
  Realize that no one is giving out Git style points. It's ok to
  ``power-cycle'', i.e.~re-initialize the Git repository, to get
  unstuck.
\end{itemize}

\subsection{Repositories and workflow}\label{repositories-and-workflow}

@@

\begin{itemize}
\tightlist
\item
  Dedicate a local directory (a.k.a ``folder'') to it.
\item
  Make it an RStudio Project. Optional but recommended; obviously only
  applies to projects involving R and users of RStudio.
\item
  Make it a Git repository.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Go about your usual business, e.g.~writing R scripts or authoring
  reports in LaTeX or R Markdown. But instead of only \emph{saving}
  individual files, periodically you make a \textbf{commit}, which takes
  a snapshot of all the files in the entire project.

  \begin{itemize}
  \tightlist
  \item
    Have you ever versioned a file
    \href{http://www.phdcomics.com/comics/archive.php?comicid=1531}{by
    adding your initials or the date}? That is effectively a
    \textbf{commit}, albeit only for a single file: it is a version that
    is significant to you and that you might want to inspect or revert
    to later.
  \end{itemize}
\item
  Push commits to GitHub periodically.

  \begin{itemize}
  \tightlist
  \item
    This is like sharing a document with colleagues on DropBox or
    sending it out as an email attachment. By pushing to GitHub, you
    make your work and all your accumulated progress accessible to
    others.
  \end{itemize}
\end{itemize}

This is a moderate change to your normal, daily workflow. It feels weird
at first, but quickly becomes second nature. In
\href{http://stat545.com}{STAT 545} students are required to submit all
coursework via GitHub, starting in week one. Most have never seen Git
before and do not identify as ``programmers''. It is a major topic in
class and office hours for the first two weeks. Then we practically
never discuss it again.

\emph{I am not saying anything about pulling and collaboration here,
which seems a bit extreme, even if I can't say much}

\subsection{Commits, diffs, and tags}\label{commits-diffs-and-tags}

@@

\begin{itemize}
\tightlist
\item
  repository
\item
  commit
\item
  diff
\end{itemize}

Recall that a repository -- or ``repo'' -- is just a directory of files
that Git is going to manage holistically. A commit functions like a
snapshot of all the files in the repo, at a specific moment. Under the
hood, that is not exactly how Git implements things. Mental models don't
have to be accurate in order to be useful, but in this case there's some
value in aligning the two.

\emph{this is really crying out for an example and/or diagrams,
something that shows commits unfolding and let's you illustrate a diff,
sort of like this from the barlett talk}

\begin{figure}
\centering
\includegraphics{bartlett-commit-history.png}
\caption{}
\end{figure}

Consider version A of a file and a modified version, version B. Assume
that version A was part of a Git commit and version B was part of the
next commit. The set of differences between A and B is called a ``diff''
and Git users contemplate diffs a lot. Diff inspection is how you
re-explain to yourself how version A differs from version B. Diff
inspection is not limited to adjacent commits. You can inspect the diffs
between any two commits.

@@

Every time you make a commit you must also write a short \emph{commit
message}. Ideally, this conveys the \emph{motivation} for the change.
Remember, the diff will show the content. When you revisit a project
after a break or need to digest recent changes made by a colleague,
looking at the \emph{history}, by reading commit messages and skimming
through diffs, is an extremely efficient way to get up to speed.

Every commit needs some sort of nickname, so you can identify it. Git
does this automatically, assigning each commit what is called a
\emph{SHA}, a seemingly random string of 40 letters and numbers (it is
not, in fact, random but is a SHA-1 checksum hash of the commit). Though
you will be exposed to these, you don't have to handle them directly
very often and, when you do, usually the first 7 characters suffice. You
can also designate certain snapshots as special with a \emph{tag}, which
is a name of your choosing. In a software project, it is typical to tag
a release with its version, e.g., ``v1.0.3''. For a manuscript or
analytical project, you might tag the version submitted to a journal or
transmitted to external collaborators.

\subsection{Markdown is special on
GitHub}\label{markdown-is-special-on-github}

@@

Any file written in Markdown is rendered in an HTML-like way on GitHub.
In particular, formatting and links ``just work''. This is the last
piece we need to seal my claim that merely pushing your project to
GitHub gives it a web presence for zero extra work. If you make even a
modest effort to embed a few explanatory Markdown files in your repo,
you will get an automatically-updated project website for free. In
particular, if a directory has a \texttt{README.md} file, GitHub renders
it like a home page or ``index.html'' when people visit that directory
in the browser. It is very common for a repo to have a top-level
\texttt{README.md}, but each subdirectory can have its own as well.

\subsection{Markdown is special for R
users}\label{markdown-is-special-for-r-users}

Markdown is extra special for R users because of R Markdown, which is
just Markdown that includes chunks of R code. Again, do not regard R
Markdown as something you must clear your schedule to learn. If you can
write email and a bit of R code, you can write R Markdown. The rmarkdown
package converts R Markdown (\texttt{.Rmd} files) to Markdown
(\texttt{.md} files), running the code and inserting the results,
including figures, into the document. This is powered by another
package, knitr, under the hood. This process is made especially easy in
RStudio, but is by no means limited to users of that application.

\emph{this too is crying out for a visual Rmd -\textgreater{} md
-\textgreater{} rendered thing looking good}

These R-derived Markdown files, if committed and pushed, then enjoy the
usual privileged treatment on GitHub already described above. Once an
\texttt{.Rmd} file has been rendered to \texttt{.md}, anyone viewing it
on GitHub can read the prose, study the R code, \textbf{and view the
results of running that code}, including figures. It is the best of all
worlds, because the code is revealed and, by definition, is the code
that produced the results. And yet a reader can gaze upon the product in
a web browser, without needing to download the code, install all
necessary dependencies, and run it.

@@

You do not, in fact, even need to work in R Markdown to exploit this. It
works with plain R scripts as well. You can use exactly the same
machinery to prepare a rendered version of an R script, i.e.~to go from
\texttt{.R} to \texttt{.md}. Again, RStudio makes this especially easy,
but this is not limited to RStudio. Once the Markdown file is pushed to
GitHub, it is as if the reader has run your code or is able to look over
your shoulder at your R session. This provides an lightweight system for
exposing work-in-progress to collaborators, without slowing down to
create separate reports. Comment lines that begin with
\texttt{\#\textquotesingle{}} are elevated to top-level prose, providing
a way to make the document more welcoming for a reader. Once there are
many prose comments, you might decide to switch from \texttt{.R} to
\texttt{.Rmd}, have proper top-level prose, and move the code down into
chunks.

\emph{figure showing the yaml needed for this? so the use of
\texttt{github\_document} or \texttt{keep\_md\ =\ TRUE}; try to make a
figure that does double or triple duty, i.e.~illustrates this and points
made elsewhere}

Let's zoom back out again and consider R Markdown and the rmarkdown
package more generally. I want to point out that R Markdown can be
rendered to many more formats than Markdown! I have emphasized the
production of Markdown here because it is extremely useful in
GitHub-hosted projects and that seems to be underappreciated. But the
rmarkdown package can convert \texttt{.Rmd} to a wide array of output
formats, including HTML, PDF, and Word (\texttt{.docx}). The Markdown
must be created regardless. Sometimes it is all you need, such as on
GitHub. In other contexts, it is just a necessary intermediate and may
even be discarded.

\subsection{Which files to commit}\label{which-files-to-commit}

@@

\begin{itemize}
\tightlist
\item
  R markdown \texttt{.Rmd} --\textgreater{} markdown \texttt{.md}
\item
  R markdown \texttt{.Rmd} --\textgreater{} markdown \texttt{.md}
  --\textgreater{} \texttt{.html} or \texttt{.pdf} or \texttt{.docx}
\item
  R script \texttt{.R} --\textgreater{} results as \texttt{.csv} or
  \texttt{.rds} and figures \texttt{.png}
\item
  LaTeX \texttt{.tex} --\textgreater{} WHAT GOES HERE NOW?
  --\textgreater{} \texttt{.pdf}
\end{itemize}

The files at the far left are clearly source files. In the case of an R
script, this is literally true, but it's morally true for R markdown and
LaTeX files too. These are files that you directly create and edit ``by
hand''.

The files at the far right are clearly derived and are often described
as ``targets''. These files are programmatically generated from source
files (and possibly other inputs). These files are the product and they
have external value, often for communicating ideas and results.

The files in the middle are intermediates. Like targets, they are
programmatically generated, but, unlike targets, no one necessarily
cares about them. However, note that intermediate Markdown \texttt{.md}
is an exception, since it is extremely useful on GitHub -- much more so
than \texttt{.html}, \texttt{.pdf}, or \texttt{.docx} -- and is more
like an additional target.

A critical issue for workflow happiness is figuring out how to manage
the production and storage of source, intermediates, and targets with
respect to Git. You can direct Git to ignore specific files or certain
types of files, such as autosaves created by your editor. This reduces
clutter in your project: Git will not pester you to commit changes to
these files and they will not appear in the associated GitHub
repository. A file that Git does not ignore is said to be
\emph{tracked}.

@@

\emph{Is it immediately useful to someone?} If so, that is a reason to
track it and push it to GitHub. There is a taboo against committing
derived products, inherited from Git's software development roots. The
reasoning is that compiled programs are platform-specific and,
therefore, people are better served by getting current source from Git
and compiling themselves. I think data analytic targets, like figures
and rendered reports, are very different beasts and it's misguided to
reflexively exclude them from version control. They are immediately
useful, especially to consumers of a project (versus the makers). To the
extent that a GitHub repo is meant for dissemination, there is no reason
to burden every consumer with unnecessary friction. Most simply will not
bother to clone the repo, install all the necessary dependencies, and
remake the products. Make them readily available.

\emph{Is it available elsewhere?} If so, then perhaps you don't need to
track it and push it to GitHub. Many people who have a policy of not
tracking derived products also have a system that makes these available
elsewhere, such as on a separate website. There are ways to automate
this via GitHub, but that is beyond the scope of this article and not
recommended for your early days with Git and GitHub. Beware those who
recommend the exclusion of derived products without offering any
practical solution for making them available some other way.

\emph{Is it huge or changing often? Is it a format that is of little use
on GitHub?} These are all good reasons to not track a file with Git.
They can make your repository bloated and slow down pushes and pulls. If
a file is binary, such as a Word document or Excel spreadsheet, Git and
Git clients will not be able to provide human-readable diffs. Neither
will GitHub be able to directly display this file in the browser. Be
aware, however, GitHub-friendliness does not just boil down to ``is it
plain text?''. GitHub has excellent support for non-code files, such as
image formats (PNG, JPG, GIF, and SVG) and PDFs. It even provides visual
diffs, which are extremely useful for understanding changes in figures.
Finally, even though HTML is plain text, it is of little direct use on
GitHub, because, unlike Markdown, it is not rendered.

\emph{Will diffs be useful to you?} Some derived files are simply too
big or miserable to read casually, such as \texttt{.csv} files of
processed results or \texttt{.html} derived from \texttt{.Rmd}. But they
may still be worth tracking with Git, because the diffs are often modest
and quite interesting. I have caught unexpected changes in analytical
results and student-facing webpages this way. When you re-run an
analysis with updated input data or after updating R packages, the diffs
presented by Git help you quickly pinpoint the downstream consequences.

\emph{Will it make collaboration harder?} \emph{talk about the scenario
Nick brought up re: binary files like PDF being a common source of merge
conflict
\url{https://github.com/dsscollection/git-github-for-stats/issues/6}}

@@

\subsection{Collaboration}\label{collaboration}

@@

\begin{itemize}
\item
  \textbf{Edit, save, attach.} In this workflow, everyone has one (or
  more!) copies of the document and they circulate via email attachment,
  accumulating initials and dates in the filename. Which one is
  ``master''? Does this question even make sense anymore? How do
  different versions of the document relate to each other? If you want
  to see a version combining the best versions of each section, how
  would you reconcile the different copies into one? All of this usually
  gets sorted out by social contract, a fairly manual process, and at
  least one miserable person.
\item
  \textbf{Google Doc.} In this workflow, there is only one copy of the
  document and it lives in the cloud. Anyone can access the most recent
  version on demand. Anyone can edit or comment or propose a change and
  this is immediately available to everyone else. Anyone can see who's
  been editing the document and how and, if disaster strikes, can revert
  to a previous version. A great deal of ambiguity and annoying
  reconciliation work has been designed away.
\end{itemize}

Managing a project via Git/GitHub is much more like the Google Doc
scenario, but also offers some of the attractive features of ``edit,
save, attach''. With Git/GitHub, collaborators can work offline and
there can be independent lines of development. The real power comes from
regular and structured reconciliation of all the files in the project
via Git/GitHub. It is definitely more complicated than collaborating on
a Google Doc, but also more powerful.
